How does the writing system get informative? Differentiation or conservation?

### 2023-07-10 manuscript/wp4.tex
ms 10 jul

\PassOptionsToPackage{hidelinks}{hyperref}
\documentclass[doc,biblatex]{apa7}
\DeclareLanguageMapping{american}{american-apa}
\addbibresource{references.bib}
\captionsetup[figure]{textfont=rm,font=small}
\usepackage{amsmath}
\usepackage{mathspec}
\setallmainfonts{Times New Roman}
\setmainfont{Times New Roman}

@@

One notable case of this is the heterographic homophones---words that sound alike but which are written with different spellings. For example, <meat> and <meet> are both pronounced /miːt/. Furthermore, these spellings can be cognitively useful . For example, faced with a sentence beginning /ðɛr/, a listener will have high uncertainty about what word---or even what sentence structure---is likely to come next: a noun, as in /ðɛr dɒg/, a form of the verb to be, as in /ðɛr ɪz/, or the progressive form of a verb, as in /ðɛr gəʊɪŋ/. In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate between these cases, allowing the reader to get directly to the meaning. This idiosyncratic spelling of words that are homophonous in speech may confer certain benefits in reading, particularly in terms of comprehension. Heterography permits a writing system to be---at least in some areas---more informative than its spoken counterpart.

@@

Is it possible, then, that heterography may be selected for in the cultural evolution of orthography? And if so, what might be the mechanism through which such selection could operate? \textcite[pp.~325--326]{Berg:2021} sketch two models of how a word might enter a state of heterography, as depicted in Fig.~\ref{berg_aronoff_models}. The first model, \textit{the differentiation model}, explains heterography through the creation of a new orthographic form. For example, in (British) English, <cheque>, which the OED describes as a ``a differentiated spelling of \textit{check},'' allows the written language to distinguish the bank order from other meanings of the word \textit{check}. The word \textit{byte} is a respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}. In English, surnames that are homophonous with ordinary nouns often add <e> (Clarke, Greene, Wilde) or double the final consonant (Carr, Hogg, Mann), and it is common for brand names to adopt creative spellings, such as <flickr> (from <flicker>), <Froot Loops> (from <fruit>), or MVMT (from <movement>).

The second model, \textit{the conservation model}, explains heterography as the historical residue of sound change: Two spoken forms merge and become homophonous, but their heterographic spellings are conserved in the orthography. For example, the words \textit{night} and \textit{knight}---heterographic homophones in modern English---were previously pronounced differently from each other (/niçt/ and /kniçt/; in Old and Middle English, /kn/ was a permissible consonant cluster, as it still is in modern Dutch and German).

Innovation

French spellings
double letters inn
add -e

@@

Participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials. In the passive exposure trials, the alien object was presented alongside the written and spoken forms for 2000~ms. The written form appeared after a 500~ms delay so that participants would first attend to the object before looking at the word. In mini-test trials, which were interleaved among the passive exposure trials, the participant had to type in the appropriate written form for one of the objects. The mini-tests were included to help participants actively engage in the learning process. After submitting a spelling, the participant was given feedback on any errors (additions shown in bold green text and deletions shown in red strikethrough text). The input box only accepted alphabetical strings of between four and nine characters, and spellings containing certain letter combinations (e.g., <bk>, <bl>, <pin>, <ye>) were not accepted to prevent participants from entering English color words or abbreviations thereof. Participants received a 1p bonus for spelling the stem correctly or a 2p bonus for spelling the entire word correctly. However, to receive the bonus, the word had to be submitted within 20~seconds. Each of the 12 object--word pairs in the training set (i.e., those that passed through the bottleneck on transmission) was passively exposed 12 times and mini-tested 3 times, resulting in a total of 144 passive exposure trials and 36 mini-test trials (the maximum possible bonus in training was therefore 72p). To ensure that participants were indeed listening to the spoken forms, a voice asked participants to click on the object at three random points during training; participants who did not follow this instruction were excluded.

@@

After training, participants assigned to the communicative conditions completed a live, over-the-internet communication game with another participant. The two participants both received training on the same orthographic system inherited from a previous generation. The communication game mirrors the overall structure of the non-communicative test with the production and comprehension trials becoming the two sides of a communicative interaction. On a given trial, one participant---the sender---would complete a production trial (produce a word for a given meaning) and this word was transmitted to the other participant. The second participant---the receiver---would then complete a comprehension trial (select an object for the transmitted word). The participants then switched roles, resulting in the same overall trial structure as the non-communicative condition (i.e., alternation between production and comprehension trials with the same bonusing scheme). However, the communication game differed from the non-communicative test in three ways. First, the transmitted and received words were displayed in text message speech bubbles to emphasize the communicative nature of the test. Second, in cases where a single wordform maps to multiple objects, the receiver performing the comprehension trial could not click any of the objects---their goal was to click the target object being communicated by the sender. Third, both participants received rich feedback: The sender saw which object the receiver clicked on, and the receiver saw which object was the correct target. Overall, then, the communicative test is identical to the non-communicative test in terms of the task to be performed, but the goal is quite different: In the non-communicative test, the goal and reward structure is based on accurately reproducing the orthography learned during training, whereas in the communicative test, the goal and reward structure is based on successfully communicating a target item.

@@

Our results suggest that the differentiation model is not a good explanation for the emergence of informative heterography.

Three pathways

- conscious design by communicators
- emergence through learning (chance alignment with color)
-

@@

Our first experiment tests the ability of the conservation model to explain the emergence of informativeness in orthography.

@@

The methods were mostly identical to Experiment~1 with three major exceptions.

sound change

non convergence

no variation

compositional start

@@

@@

Each iterated learning chain was run for 12 generations, divided into four epochs of three generations. In the first epoch, all suffixes were distinct, allowing the spoken language to express all four colors, but by the fourth epoch all suffixes were reduced into a single phonetic form, making the spoken language entirely uninformative about color. This was achieved through three sound changes occurring between the four epochs. In the first sound change, two of the suffixes were chosen at random to be merged. To perform the merger, the consonant from one of the suffixes (chosen at random) was paired with the vowel from the other suffix, resulting in a new suffix form that would replace the original two. For example, assuming the language has the suffixes /ɪθə/ (pink), /ɪsɛɪ/ (gray), /ɪʃəʊ/ (blue), and /ɪhu/ (yellow), we might merge /ɪsɛɪ/ and /ɪhu/ into /ɪhɛɪ/, such that the spoken language is no longer able to mark a distinction between gray and yellow. In the second sound change, the same procedure is applied to the two suffixes that were not selected previously (e.g., /ɪθə/ and /ɪʃəʊ/ become /ɪθəʊ/). And in the third and final sound change, the two suffix forms created through the first and second sounds changes---/ɪhɛɪ/ (grey/yellow) and /ɪθəʊ/ (pink/blue)---merge into a single form (e.g., /ɪθɛɪ/). Importantly, the spellings do not change automatically following a sound change event; instead, the orthographic system is free to adapt (or not) in response to sound changes.

>>>>

\PassOptionsToPackage{hidelinks}{hyperref}
\documentclass[doc,biblatex]{apa7}
\DeclareLanguageMapping{american}{american-apa}
\addbibresource{references.bib}
\captionsetup[figure]{textfont=rm,font=small,labelfont=bf}
\usepackage{amsmath}
\usepackage{mathspec}
\setallmainfonts{Times New Roman}
\setmainfont{Times New Roman}
% \setsansfont{Helvetica Neue}

@@

One notable case of this is the heterographic homophones---words that sound alike but which are written with different spellings. For example, <meat> and <meet> are both pronounced /miːt/. Furthermore, these spellings can be cognitively useful . For example, faced with a sentence beginning /ðɛr/, a listener will have high uncertainty about what word---or even what sentence structure---is likely to come next: a noun, as in /ðɛr dɒg/, a form of the verb to be, as in /ðɛr ɪz/, or the progressive form of a verb, as in /ðɛr gəʊɪŋ/. In writing, by contrast, this uncertainty is greatly reduced; the spellings <their>, <there>, and <they're> differentiate between these cases, giving the reader a headstart on parsing the upcoming syntax. This idiosyncratic spelling of words that are homophonous in speech may confer certain benefits in reading, particularly in terms of comprehension. Heterography permits a writing system to be---at least in some areas---more informative than its spoken counterpart.

@@

Is it possible, then, that heterography may be selected for in the cultural evolution of orthography? And if so, what might be the mechanism through which such selection could operate? \textcite[pp.~325--326]{Berg:2021} sketch two models of how a word might enter a state of heterography, as depicted in Fig.~\ref{models_of_heterography}. The first model, \textit{the differentiation model}, explains heterography through the creation of a new orthographic form. For example, many monosyllabic words that are homophonous with function words have tended to adopt alternate spellings (usually by adding material): \textit{be}--\textit{bee}, \textit{but}--\textit{butt}, \textit{by}--\textit{bye}--\textit{buy}, \textit{for}--\textit{fore}--\textit{four}, \textit{in}--\textit{inn}, \textit{or}--\textit{oar}--\textit{ore}, \textit{so}--\textit{sew}, \textit{to}--\textit{too}--\textit{two}, etc. Other cases of heterography appear to be the result of a very conscious choice to disambiguate meanings; for example, in British English, the word \textit{cheque} was introduced to to distinguish the bank order from other meanings of the word \textit{check}, the word \textit{byte} is a deliberate respelling of \textit{bite} to avoid accidental mutation into the closely related term \textit{bit} \parencite{Buchholz:1977}, and the word \textit{lite} is a respelling of \textit{light} that is frequently used in food products to distinguish light-in-weight from light-in-color. Etymological (and folk-etymological) spellings introduced during the Renaissance also sometimes resulted in heterographic homophones, perhaps contributing to their success: \textit{scene}--\textit{seen}, \textit{scent}--\textit{sent}, \textit{whole}--\textit{hole} \parencite[pp.~58--59]{Scragg:1974}. Heterography is also very common among proper nouns: Surnames that are homophonous with ordinary nouns often add <e> (\textit{Clarke}, \textit{Greene}, \textit{Wilde}) or double the final consonant (\textit{Carr}, \textit{Hogg}, \textit{Mann}), and it is common for brand names to adopt creative spellings, such as \textit{flickr}, \textit{Froot Loops}, or \textit{MVMT} (movement).

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/models_of_heterography.eps}}
	\vspace*{2pt}
	\caption{Two models of heterography. In the differentiation model, two meanings (represented by pink and blue) are, at time $T_1$, expressed by a single phonetic form $P_1$ and a single orthographic form $O_1$, but by time $T_2$, two orthographic forms have emerged to differentiate the meanings in writing. In the conservation model, the two distinct phonetic forms that existed at time $T_1$ have become homophonous by time $T_2$, but the two corresponding orthographic forms have been conserved, resulting in the same state of heterography as in the differentiation model. Adapted from \textcite[pp.~325--326]{Berg:2021} with permission.}
	\label{models_of_heterography}
	\end{figure}

The second model, \textit{the conservation model}, explains heterography as the historical residue of sound change: Two (or more) spoken forms merge and become homophonous, but the original heterographic spellings are conserved in the orthography. In English, for example, the meat--meet merger ultimately resulted in Middle English /ɛ:/ (spelled <ea>) and /e:/ (spelled <ee>) merging into /i:/ in Early Modern English, although the spellings were never updated accordingly, thus giving rise to a set heterographic homophones that persist in present-day English: \textit{heal}--\textit{heel}, \textit{leak}--\textit{leek}, \textit{meat}--\textit{meet}, \textit{read}--\textit{reed}, \textit{sea}--\textit{see}, \textit{team}--\textit{teem}, \textit{weak}--\textit{week}, etc. Similarly, other historic sound changes have likely resulted in (or contributed to) pairs of words entering a state of heterography, such as the reduction of /kn/ into /n/ (\textit{knight}--\textit{night}, \textit{know}--\textit{no}, \textit{knot}--\textit{not}, etc.), the loss of /ç/ (\textit{eight}--\textit{ate}, \textit{right}--\textit{rite}, \textit{sight}--\textit{site}, etc.), and the merger of /ʍ/ into /w/ (\textit{whale}--\textit{wail}, \textit{which}--\textit{witch}, \textit{whine}--\textit{wine}, etc.).

@@

Participants were trained on the spoken and written forms through a combination of passive exposure trials and ``mini-test'' trials. In the passive exposure trials, the alien object was presented alongside the written and spoken forms for 2000~ms. The written form appeared after a 500~ms delay so that participants would first attend to the object before looking at the word. In mini-test trials, which were interleaved among the passive exposure trials, the participant had to type in the appropriate written form for one of the objects. The mini-tests were included to help participants actively engage in the learning process. After submitting a spelling, the participant was given feedback on any errors (additions shown in bold green text and deletions shown in red strikethrough text). The input box only accepted lowercase alphabetical strings of between four and nine characters, and spellings containing certain letter combinations (e.g., <bk>, <bl>, <pin>, <ye>) were not accepted to prevent participants from entering English color words or abbreviations thereof. Participants received a 1p bonus for spelling the stem correctly or a 2p bonus for spelling the entire word correctly. However, to receive the bonus, their response had to be submitted within 20~seconds. Each of the 12 object--word pairs in the training set (i.e., those that passed through the bottleneck on transmission) was passively exposed 12 times and mini-tested 3 times, resulting in a total of 144 passive exposure trials and 36 mini-test trials (the maximum possible bonus in training was therefore 72p). To ensure that participants were indeed listening to the spoken forms, a voice asked participants to click on the object at three random points during training; participants who did not follow this instruction were excluded.

@@

After training, participants assigned to the communicative conditions completed a live, over-the-internet communication game with another participant. The two participants both received training on the same orthographic system inherited from a previous generation, albeit subject to independent bottlenecks on transmission. The communication game mirrors the overall structure of the non-communicative test with the production and comprehension trials becoming the two sides of a communicative interaction. On a given trial, one participant---the sender---would complete a production trial (produce a word for a given meaning) and this word was transmitted to the other participant. The second participant---the receiver---would then complete a comprehension trial (select an object for the received word). The participants then switched roles, resulting in the same overall trial structure as the non-communicative condition (i.e., alternation between production and comprehension trials with the same bonusing scheme). However, the communication game differed from the non-communicative test in three ways. First, the transmitted and received words were displayed in text message speech bubbles to emphasize the communicative nature of the test. Second, in cases where a single wordform maps to multiple objects, the receiver performing the comprehension trial could not click any of the objects---their goal was to click the target object being communicated by the sender. Third, both participants received rich feedback: The sender saw which object the receiver clicked on, and the receiver saw which object was the correct target. Overall, then, the communicative test is identical to the non-communicative test in terms of the task to be performed, but the goal is quite different: In the non-communicative test, the goal and reward structure is based on accurately reproducing the orthography learned during training, whereas in the communicative test, the goal and reward structure is based on successfully communicating a target item.

@@

Our first experiment tested the ability of the differentiation model to explain the emergence of an informative, heterographic orthography, but we found little evidence of systematic orthographic differentiation. Exposed to an orthography with high variation, learners could have gradually conditioned that variation on the dimension where it would have been useful from the point of view of disambiguation, color, yielding differentiated forms in the orthography. But this did not happen, even under the communicative pressure. Instead, when spelling variation was conditioned on anything, it was conditioned on shape, resulting in redundant suffix spellings. This result is in stark contrast to most prior experimental iterated learning studies, in which informative, compositional systems do typically emerge, especially under communicative pressure. So, what was different about the present experiment? The primary difference was the presence of a spoken language that is decidedly \textit{not} informative, which acts as a cue to participants that the language does not mark color and that, therefore, the orthography should not mark color. Faced with variation that could be conditioned on shape or color, learners rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language. As a result, variation in spelling comes to be associated with particular stems, resulting in the emergence of idiosyncratic suffix spellings that serve no real purpose. A rough analog of this redundancy can be found in English in the spelling of the /ʃən/ suffix (<cian>, <sion>, <ssion>, or <tion>), which is largely conditioned on the stem (e.g., \textit{magician}, \textit{expansion} \textit{transmission}, and \textit{invention}).

Another pathway through which differentiation could have occurred is through conscious innovation, especially in the communicative conditions, where participants could have increased their bonuses substantially by innovating a system of color markers. However, there was no evidence of such innovation, and only four participants mentioned that they even \textit{considered} this possibility in their post-test comments (e.g., ``I wondered if I should use the color suffix rule even where I did not think it was the right word to try and help my partner get the right color object''). Why do participants choose not to innovate or even regularize an imperfect system? Firstly, although innovation was not expressly forbidden, it was perhaps suggested by the overall structure of the experiment that we expected them to use the orthography learned during training. Participants who adopted this interpretation may have feared diverging from their training either because they themselves wanted to avoid breaking the rules or because they did not know whether their partner would feel comfortable breaking the rules. Secondly, even if participants are willing to diverge, they are presented with an alignment problem: How can they decide what system of color marking to use when there is no channel through which they can have a meta-discussion? They can only use trial-and-error, and even then, the process is complicated by participants simultaneously adapting to each other's systems. Interestingly, these are dynamics that also exist in the real world, where deviation from received, sacrosanct spellings is considered socially unacceptable or uneducated (e.g., the non-standard spelling <thru> has struggled for decades---or perhaps centuries---to make any headway against <through>), and even under more progressive attitudes to change and innovation, the problem of achieving community buy-in remains, especially when the change is not imposed by top-down authority: If I decided to start using the spelling <banque> to differentiate the financial institution from river banks, would anyone take my spelling seriously or even understand the distinction I was trying to make?

Overall, our first experiment suggests that differentiation---be it through a self-organizing process in learning or through a conscious process of innovation in use---is not a good model of heterography, unless enforced by top-down authority. We now turn to consider the conservation model.

@@

Our second experiment investigates the conservation model: Given than an informative system already exists (both in speech and in writing), will the informative system persist even after the degeneration of the spoken language towards homophony? And does this happen preferentially in the presence of a shared communicative task. How much homophony can an informative, heterographic system resist under an increasingly strong simplicity pressure.

@@

The methods were identical to Experiment~1 with the following major exceptions. Firstly, and most notably, the artificial language used to seed each chain starts out fully compositional (in both its spoken and written forms), since we are now interested in whether the orthographies will maintain informativeness rather than gain it. Secondly, and consequently, the convergence-based approach used in Experiment~1 to determine the number of generations is no longer appropriate, so we now use a fixed number of generations. Thirdly, we simulate regular sound mergers which result in the spoken forms of the suffixes becoming increasingly less informative over generational time (this contrasts with Experiment~1, where the spoken form of the suffixes were always fully uninformative).

@@

We recruited xxx participants via the Prolific platform. The payment and bonusing scheme was identical to Experiment~1 (median bonus: £0.xxx). The median completion time was xxx:xxx, resulting in an hourly rate of £xxx (£xxx including bonus). Unlike Experiment~1 we restricted recruitment to native speakers of English because it was particularly important that participants would be able to distinguish the voiceless fricatives /θ s ʃ h/ and we found in piloting that non-native speakers of English frequently struggled with these distinctions. xxx participants were rejected because they did not pass the random audio checks, and a further xxx participants were removed because they had been partnered with one of the rejected participants. In addition, xxx participants were lost to communication-game pairing failures, resulting in a final dataset of 360 participants (organized into ten chains of twelve generations in each condition, with two participants per generation in Learning \& Communication).

@@

Each iterated learning chain was run for 12 generations, divided into four epochs of three generations. In the first epoch, all spoken suffixes were distinct, allowing the spoken language to express all four colors, but by the fourth epoch all suffixes were reduced into a single phonetic form, making the spoken language entirely uninformative about color. This was achieved through three sound changes occurring between the four epochs. In the first sound change, two of the suffixes were chosen at random to be merged. To perform the merger, the consonant from one of the suffixes (chosen at random) was paired with the vowel from the other suffix, resulting in a new suffix form that would replace the original two. For example, assuming the language has the suffixes /ɪθə/ (pink), /ɪsɛɪ/ (gray), /ɪʃəʊ/ (blue), and /ɪhu/ (yellow), we might merge /ɪsɛɪ/ and /ɪhu/ into /ɪhɛɪ/, such that the spoken language is no longer able to mark a distinction between gray and yellow. In the second sound change, the same procedure is applied to the two suffixes that were not selected previously (e.g., /ɪθə/ and /ɪʃəʊ/ become /ɪθəʊ/). And in the third and final sound change, the two suffix forms created through the first and second sounds changes---/ɪhɛɪ/ (grey/yellow) and /ɪθəʊ/ (pink/blue)---merge into a single form (e.g., /ɪθɛɪ/). Importantly, the spellings do not change automatically following a sound change event; instead, the orthographic system is free to adapt (or not) in response to sound changes.
